Non-word repetition in Yélî Dnye

### 2020-12-21 manuscript.Rmd
add item complexity disc subsection draft

Next, we investigated patterns of deletion and substitution. Deletions were relatively rare: there were only `r sum(del_tab[c(oralVs,nasalVs)])` instances of deleted vowels (~`r mean(del_pc[c(oralVs,nasalVs)],na.rm=T)`% of all vowel targets), and `r sum(del_tab[c(simplexCs,complexCs)])` instances of deleted consonants (~`r mean(del_pc[c(simplexCs,complexCs)],na.rm=T)`% of all consonant targets). Substitutions were somewhat more frequent, but still relatively rare. Nasal vowels were produced as oral vowels and vice versa (`r sum(sub_tab[oralVs,nasalVs])` oral targets as nasal (~`r mean(sub_pc[oralVs,nasalVs],na.rm=T)`% of targets) and `r sum(sub_tab[nasalVs,oralVs])` nasal targets produced as oral (~`r mean(sub_pc[nasalVs,oralVs],na.rm=T)`% of targets). Both oral and nasal vowels were also occasionally subsituted by vowels of a different quality that preserved the target oral/nature airflow (`r sum(sub_tab[oralVs,oralVs])` oral vowels (~`r mean(sub_pc[oralVs,oralVs],na.rm=T)`% of targets and `r sum(sub_tab[nasalVs,nasalVs])` nasal vowels (~`r mean(sub_pc[nasalVs,nasalVs],na.rm=T)`% of targets). Among consonants, complex consonants were occasionally substituted by simpler ones (specifically `r print.phone.list(complexCs.ipa)` as `r print.phone.list(simplexCs.ipa)`), `r sum(sub_tab[complexCs,simplexCs])` times, in approximately `r mean(sub_pc[complexCs,simplexCs],na.rm=T)`% of all complex consonant targets. Interestingly, simple consonants were sometimes also substituted by complex ones; this happened `r sum(sub_tab[simplexCs,complexCs])` times, in approximately `r mean(sub_pc[complexCs,simplexCs],na.rm=T)`% of all simple consonant targets. Finally, simple consonants were mispronounced as other simple consonants (`r sum(sub_tab[simplexCs,simplexCs])` instances, ~`r mean(sub_pc[simplexCs,simplexCs],na.rm=T)`% of all simple consonant targets) and complex consonants as other complex consonants (`r sum(sub_tab[complexCs,complexCs])` instances, ~`r mean(sub_pc[complexCs,complexCs],na.rm=T)`% of all complex consonant targets). Note that the substitutions included phones that are not native to Yélî Dnye but do occur in English (e.g., [tʃ]); as these data come from careful transcriptions by a native Yélî Dnye speaker who is very fluent in English, we take these segmental substitutions faithfully as an indication that several of our participants have mastered production of some English phones, possibly produced within whole English word forms (see below).

@@

- What is the overall repetition scores (whole word, phoneme based, distance)?

- How does this change as a function of item complexity (number of syllables, sound complexity)?

1. Prediction based on previous work: Children have higher scores for shorter items
1. Prediction for Yélî made before seeing the data: The length distribution in Yélî words is more balanced than that in English, and thus the score decline for poly- versus mono-syllables may be less pronounced than that for English. 
1. it turns out we were right! the scores for 2-4 syll words decline only slightly with length

2. Prediction based on previous work: Similarly, we do not know of NWR research that manipulates the difficulty of the sounds that are included in the items, but word naming and other research suggests that children achieve higher scores when producing easy and/or typologically common sounds than difficult and/or typologically rare sounds [CITE]. Therefore, we expect higher scores for items with common sounds than in those with rare sounds.
2. Prediction for Yélî made before seeing the data: The Yélî sound inventory is very large and compressed, with many similar sounds that are acoustic and articulatory neighbors. Therefore, this may constitute a pressure for children to have finer auditory skills (and perhaps more precise articulations) than children speaking languages with a simpler inventory. As a result, differences between easier and harder items may be smaller in this work than in other research. 
2. it turns out we do see

- How frequent are errors that result in real words? Is that a function of item complexity?
- Is individual variation explainable by child age, sex, birth order, monolingual status, and/or parental education?

>>>>

Next, we investigated patterns of deletion and substitution. Deletions were relatively rare: there were only `r sum(del_tab[c(oralVs,nasalVs)])` instances of deleted vowels (~`r mean(del_pc[c(oralVs,nasalVs)],na.rm=T)`% of all vowel targets), and `r sum(del_tab[c(simplexCs,complexCs)])` instances of deleted consonants (~`r mean(del_pc[c(simplexCs,complexCs)],na.rm=T)`% of all consonant targets). Substitutions were somewhat more frequent, but still relatively rare. Nasal vowels were produced as oral vowels and vice versa (`r sum(sub_tab[oralVs,nasalVs])` oral targets as nasal (~`r mean(sub_pc[oralVs,nasalVs],na.rm=T)`% of targets) and `r sum(sub_tab[nasalVs,oralVs])` nasal targets produced as oral (~`r mean(sub_pc[nasalVs,oralVs],na.rm=T)`% of targets)). Both oral and nasal vowels were also occasionally subsituted by vowels of a different quality that preserved the target oral/nature airflow (`r sum(sub_tab[oralVs,oralVs])` oral vowels (~`r mean(sub_pc[oralVs,oralVs],na.rm=T)`% of targets and `r sum(sub_tab[nasalVs,nasalVs])` nasal vowels (~`r mean(sub_pc[nasalVs,nasalVs],na.rm=T)`% of targets). Among consonants, complex consonants were occasionally substituted by simpler ones (specifically `r print.phone.list(complexCs.ipa)` as `r print.phone.list(simplexCs.ipa)`), `r sum(sub_tab[complexCs,simplexCs])` times, in approximately `r mean(sub_pc[complexCs,simplexCs],na.rm=T)`% of all complex consonant targets. Interestingly, simple consonants were sometimes also substituted by complex ones; this happened `r sum(sub_tab[simplexCs,complexCs])` times, in approximately `r mean(sub_pc[complexCs,simplexCs],na.rm=T)`% of all simple consonant targets. Finally, simple consonants were mispronounced as other simple consonants (`r sum(sub_tab[simplexCs,simplexCs])` instances, ~`r mean(sub_pc[simplexCs,simplexCs],na.rm=T)`% of all simple consonant targets) and complex consonants as other complex consonants (`r sum(sub_tab[complexCs,complexCs])` instances, ~`r mean(sub_pc[complexCs,complexCs],na.rm=T)`% of all complex consonant targets). Note that the substitutions included phones that are not native to Yélî Dnye but do occur in English (e.g., [tʃ]); as these data come from careful transcriptions by a native Yélî Dnye speaker who is very fluent in English, we take these segmental substitutions faithfully as an indication that several of our participants have mastered production of some English phones, possibly produced within whole English word forms (see below).

@@

We used non-word repetition to investigate phonological development in a language with a large phonological inventory (including some typologically rare segments). While the study, in itself, advances NWR research by demonstrating the successful application of this method in a rural, non-industrialized population [see also @cristia2020infant], the results also inform the larger body of NWR work regarding the influence of stimulus length, segmental frequency, age, and developmental context, where previous findings have often been mixed.

In our sample of `r nrow(overall)` children between ages `r min(overall$Age)` and `r max(overall$Age)`, the average item-level score was `r round(mean(overall$Word))`%, falling well within the expected range of NWR applications in previous studies on a variety of linguistic populations (see Figure \@ref(fig:prevlit-fig)). Segment-level analyses revealed that repetitions recovered an average of `r round(mean(overall$Phoneme))`% of the segments in the original stimulus, indicating that inexact repetitions often constituted _mostly_ accurate reroductions of the target. Complementary data from segment-based normalized Levenshtein distance reflect this same pattern, with an average of `r round(mean(overall$NLD))`% of the stimulus segments getting substituted, inserted, or deleted upon repetition. Paired with our thorough training protocol, we take these NWR scores as indicating that (a) our adaptations to NWR for this context were successful, even given a number of non-standard changes to the training phase and to the design of the stimuli and (b) Yélî children show comparable performance to others tested on a similar task, despite the many linguistic, cultural, and socioeconomic differences between this and previously tested populations. Given successful (and comparable) execution of this task, we can dive deeper into effects of stimulus length, segmental frequency, age, and developmental context.

### Item complexity

We investigated the effect of item complexity on NWR scores by varying both the number of syllables in the item and its average segmental frequency. Based on previous work, we had predicted that children would have higher NWR scores for shorter items. That said, previous work has shown both very small [@piazzalunga2019articulatory] and very large [@cristia2020infant; @jabere2018xperiment] effects of stimulus length and, further, the Yélî Dnye dictionary suggests that mono- and bi-syllabic words are nearly equally frequent in the current language, with trisyllabic and longer words making up a non-trivial 10% of the remaining words. Compare this to, for example, English, which is substantially more skewed toward monosyllabic word forms **Alex I'm going off your note here ("Prediction for Yélî made before seeing the data: The length distribution in Yélî words is more balanced than that in English, and thus the score decline for poly- versus mono-syllables may be less pronounced than that for English.""). I don't have a reference for this, can you please finish the thought or nix this bit?**. Setting aside our monosyllabic stimuli, which all contained typologically infrequent segments, we can examine effects of item length among the remaining stimuli, which range between 2 and 4 syllables long. While indeed NWR scores were overall lower for longer items (e.g., see Figure \@ref(fig:prevlit-fig)), the effect of item length was not significant in a statistical model that additionally accounted for age and random effects of item and participant. In light of mixed prior results of item length, we propose two possible (and non-mutually exclusive) explanations for this minimal impact of item length. First, further extensions of this type of analysis in more populations may reveal that, in general (and cross-linguistically), item length effects are variable between languages, potentially reflecting the distribution of word lengths in the ambient language and other (morpho-)phonological tendencies in the lexicon. Second, above and beyond these language-specific effects, the general impact of item length on NWR score may be relatively small, as shown in Piazzalunga et al.'s [-@piazzalunga2019articulatory] study on Italian and as borne out in the current dataset once controlling for other factors.

Our monosyllabic items included typologically rare segments so that we could test whether lower average segmental frequency is associated with lower NWR scores. Typologically common sounds are associated with higher performance on a handful of other tasks (**REFS -- M2A: Alex, I added this based on your note, where it sounded like you had some particular studies in mind?**) though to our knowledge this has not yet been tested with non-word repetition. Regarding Yélî Dnye in particular, the phonemic inventory is both large and acoustically packed, in addition to containing several typologically infrequent (or unique) contrasts. We therefore expected to see that, while NWR scores would be lower for stimuli with lower average frequency, this effect would be relatively weak because the ambient language puts pressure on Yélî children to distinguish (perceptually and articulatorily) fine-grained phonetic differences in order to successfully communicate with others. Indeed we found a robust effect of average segmental frequency on NWR performance: Even accounting for age and random effects of item and participant, we see that target words with more frequent segments were repeated correctly more often. This effect is large, with a magnitude more than twice the size of the effect of participant age. This significant effect remains even once also accounting for the frequencies of these segments in Yélî Dnye child-directed speech, which are correlated with their typological frequencies. In sum, typological frequency effects, which have been found in other measurements of phonological processing. appear to strongly affect NWR performance, and do not appear mitigated by language-specific pressure to make finer-grained differences earlier in development.

With respect to the types of errors in repetition made, we did not see clear patterns to further guide our discussion: base rates of deletion and substitution were fairly low and the relative distribution of errors over, e.g., nasal vs. oral vowels and simple vs. complex consonants, revealed no remarkable bias in error types. That said, the lack of a difference could be due to relative imbalance across our stimuli in the use of these phonemic features (e.g., we included many more more oral than nasal targets) and future work should investigate such sources of error bias more systemtically. Some portion of the errors were introduced when the participant produced a real word (in Yélî Dnye or English) in response to the stimulus. Real-word repetitions here made up two thirds of errorful repetitions---this is quite high compared to past work [e.g., @castro1998], but it is unclear what caused this pattern in the current study: Castro and colleagues' [-@castro1998] study focused on adults rather than children, the task was administered by a team including a foreign, English-speaking researcher, and the particularities of the Yélî Dnye phonological inventory result in many true-word phonetic neighbors. Follow-up work exploring this type of error in children from other populations in addition to further work on Yélî children will clarify this effect.

### Individual differences

- Is individual variation explainable by child age, sex, birth order, monolingual status, and/or parental education?
